A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2015-10-05 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Commit Annika's corrections

% \institute{
%   Helsinki, Finland\\
%   \email{annika.siders@helsinki.fi}
%   \and 
%   Vienna University of Technology, Austria ; Australian National University \\
%   \email{bruno.wp@gmail.com} 
% }

@@

\begin{abstract}
This paper presents two detailed formalizations of ontological arguments in a simple natural deduction calculus. The first formal proof closely follows the hints in Scott's manuscript about G\"odel's argument and fills its gaps, thus verifying its correctness. The second formal proof improves the first one, by: firstly, relying on the weaker modal logic {\bf KB} instead of the stronger modal logic {\bf S5} in the proof of the last theorem;  and, secondly, not using equality to prove the first theorem. The second proof is also technically shorter than the first one, because it avoids unnecessary detours and uses Axiom 1 only once.
\end{abstract}

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considers a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence. St. Anselm's argument was further elaborated by Descartes, Leibniz and Kant.

@@

G\"odel bulit on Leibniz's work \cite{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \cite{Goedel} as well as notes by Scott \cite{scott}, in whom he confided his proof.

The increase in formality of the ontological argument has required a development of its basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being god-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a god-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{scott}, this axiom of G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

@@

The natural deduction style was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of proof calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, natural deduction formalizations can be verified step-by-step in such proof assistants, and we have in fact done this.

@@

Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to an axiom or theorem with proof shown elsewhere. When proof trees are too large to fit in the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

\begin{calculus}
{propositional rules}
{fig:PropositionalRules}

@@

\begin{calculus}
{double negation elimination}
{fig:Classical}
\infer[\neg\neg_E]{A}{ \neg\neg A }
\end{calculus}

\begin{calculus}
{quantifier rules}
{fig:QuantifierRules}

@@

\begin{calculus}
{Rules for Modal Operators}
{fig:NDK}

@@

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}. The formula $\all \psi. (\nec \psi \imp \pos \psi)$, for example, is not valid in \textbf{K} but would be provable according to the following proof:

@@

The example proof above is unsound according to the {\NDK} calculus, because the eigen-box condition is violated: the box labelled by $\omega$ is not accessed by any strong inference.

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible worlds} semantics, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \cite{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \cite{hajek}). Nevertheless, some authors of variants \cite[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

In this section, natural deduction proofs of a three convenient derived modal principles are presented. These principles are used in the ontological proofs formalized in the next sections. Although they are well-known among modal logicians, we include their proofs here in order to have a self-contained and fully formal presentation of the ontological proofs, without any gaps.

@@

\noindent
Below we exhibit a natural deduction proof for the first principle listed above  when $n = 1$. Informally, this particular case can be read as the claim that anything that is possibly necessary is in fact necessary, and it is sufficient for the ontological proofs of G\"odel and Scott. The proof of this principle depends on the previously proven Brouwer's reduction principle, on the Modal Axiom 5, and on K, whose derivation is shown in the proof of Theorem \ref{theorem:completeness}.

@@

In this section we present a detailed formalization of Scott's proof in the natural deduction calculus defined in Section \ref{sec:calculus}. The reasoning in Scott's manuscript was reproduced step-by-step and any reasoning gaps were filled in ways that we considered most natural. All derived formulas that do appear in Scott's manuscript are marked with a $\dagger$ here. Unmarked formulas were derived in the process of filling the gaps between the marked formulas mentioned in Scott's manuscript.

@@

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', when in fact it is just a convenient abbreviation.

Besides the 5 axioms and 3 definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but it gives a shorter proof.

@@

\begin{definition}
\label{D1}
A \emph{god-like} being possesses all positive properties:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
$$
\end{definition}

@@

\begin{theorem}
\label{T2}
Being god-like is an essence of any god-like being:
$$
\all x. G(x) \imp \ess{G}{x}
$$
\end{theorem}
\begin{proof}\hfill

@@

\begin{lemma}
\label{L1}
If there is a god-like being, then there is a god-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{lemma}
\label{L2}
If the existence of a god-like being is possible, then it is necessary:
$$
\pos \ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{theorem}
\label{T3}
Necessarily, there exists a god-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

Comparing this new proof of Theorem 1 with the proof of Theorem 1 indicated by G\"odel and Scott, it is noticeable that new proof is significantly shorter. The key idea to obtain a shorter proof is to instantiate the second universally quantified property of Axiom 2 by $\lambda x. \neg \rho(x)$ instead of using $=$ or $\neq$ as G\"odel and Scott did. This eliminates the reliance of the ontological argument on the presence of $=$ and $\neq$ in the logical language. Furthermore, Axiom 2 is then needed only once.

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly and informally described by Anderson \cite[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may only wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, if this suggests that there might have been a change (perhaps evolution) in the way logicians construct formal proofs; or even from a philosophical point of view, if G\"odel's preference was philosophically motivated.

@@

\setcounter{theorem}{2}
\begin{theorem}
\label{T3}
Necessarily, there exists a god-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

Since the iteration principle requires the strong modal logic {\bf S5} while Browser's reduction principle relies only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{iteration principle}.
Anderson \cite[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep{sobel2}[p. 152] acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in Section \ref{sec:newproof} relies on {\bf KB}, which is weaker than {\bf B}.

@@

If G\"odel's ontological proof is abstractly analyzed, then we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being God-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.

@@

\bibitem[{\itshape Church (1940)}]{church}
Church, A. 1940. {\itshape A Formulation of the Simple Theory of Types}, Journal of Symbolic Logic, 5: 56?68.

@@

\bibitem[{\itshape Hajek (1996)}]{hajek}
Hajek, P. 1996. Magari and others on G\"odel's Ontological Proof, in Ursini et alii
(eds.) Logic and Logical Algebra, pp. 125-136, Marcel Dekker.

@@

\bibitem[{\itshape Sobel (1987)}]{sobel}
Sobel, J. H. 1987. {\itshape G\"odel's Ontological Proof}. In: J. J. Thompson (ed.). {\itshape On being and saying : essays for Richard Cartwright},  MIT Press.

>>>>

% \institute{
%   University of Helsinki, Finland\\
%   \email{annika.siders@helsinki.fi}
%   \and 
%   Vienna University of Technology, Austria ; Australian National University \\
%   \email{bruno.wp@gmail.com} 
% }

@@

\begin{abstract}
This paper presents two detailed formalizations of ontological arguments in a simple natural deduction calculus. The first formal proof closely follows the hints in Scott's manuscript about G\"odel's argument and fills in the gaps, thus verifying its correctness. The second formal proof improves the first one, by: relying on the weaker modal logic {\bf KB} instead of the stronger modal logic {\bf S5};  and by avoiding the introduction of the equality predicate. The second proof is also technically shorter than the first one, because it eliminates unnecessary detours and uses Axiom 1 for the positivity of properties only once.
\end{abstract}

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considers a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also had the property of existence. St. Anselm's argument was further elaborated by Descartes, Leibniz and Kant.

@@

G\"odel bulit on Leibniz's work \cite{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \cite{Goedel} as well as in notes by Scott \cite{scott}, in whom he confided his proof.

The increase in formality of the ontological argument has required a development of its basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being God-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a God-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{scott}, this axiom o!
 f G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

@@

The natural deduction style was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of proof calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiar with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, natural deduction formalizations can be verified step-by-step in such proof assistants, and we have in fact done this.

@@

Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to an axiom or theorem with the proof shown elsewhere. When proof trees are too large to fit on the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

\begin{calculus}
{Propositional rules}
{fig:PropositionalRules}

@@

\begin{calculus}
{Double negation elimination}
{fig:Classical}
\infer[\neg\neg_E]{A}{ \neg\neg A }
\end{calculus}

\begin{calculus}
{Quantifier rules}
{fig:QuantifierRules}

@@

\begin{calculus}
{Rules for modal operators}
{fig:NDK}

@@

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}. For example, the formula $\all \psi. (\nec \psi \imp \pos \psi)$ is not valid in \textbf{K} but would be provable without this restriction:

@@

This example proof is unsound according to the {\NDK} calculus, because the eigen-box condition is violated: the box labelled by $\omega$ is not accessed by any strong inference.

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \cite{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \cite{hajek}). Nevertheless, some authors of variants \cite[footnotes !
 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

In this section, natural deduction proofs of three convenient derived modal principles are presented. These principles are used in the ontological proofs formalized in the next sections. Although they are well-known among modal logicians, we include their proofs here in order to have a self-contained and fully formal presentation of the ontological proofs, without any gaps.

@@

\noindent
Below we exhibit a natural deduction proof for the first principle listed above  when $n = 1$. Informally, this particular case can be read as the claim that anything that is possibly necessary is in fact necessary, and it is sufficient for the ontological proofs of G\"odel and Scott. The proof of this principle depends on the previously proven Brouwer's reduction principle, on the Modal Axiom 5, and on K, which was shown to be derivable in the proof of Theorem \ref{theorem:completeness}.

@@

In this section we present a detailed formalization of Scott's proof in the natural deduction calculus defined in Section \ref{sec:calculus}. The reasoning in Scott's manuscript has been reproduced step-by-step and all reasoning gaps are filled in by using the deduction considered most natural. All derived formulas that do appear in Scott's manuscript are marked with a $\dagger$ here. Unmarked formulas were derived in the process of filling the gaps between the marked formulas mentioned in Scott's manuscript.

@@

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', w!
 hen in fact it is just a convenient abbreviation.

Beside the 5 axioms and 3 definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but allows a shorter proof.

@@

@@

\begin{theorem}
\label{T2}
Being god-like is an essence of any God-like being:
$$
\all x. G(x) \imp \ess{G}{x}
$$
\end{theorem}
\begin{proof}\hfill

@@

\begin{lemma}
\label{L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{lemma}
\label{L2}
If the existence of a God-like being is possible, then it is necessary:
$$
\pos \ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

Comparing this new proof of Theorem 1 with the proof of Theorem 1 indicated by G\"odel and Scott, it is noticeable that new proof is significantly shorter. The key idea to obtain a shorter proof is to instantiate the second universally quantified property of Axiom 2 by $\lambda x. \neg \rho(x)$ instead of using the predicate symbols $=$ or $\neq$ as G\"odel and Scott did. This eliminates the reliance of the ontological argument on the presence of $=$ and $\neq$ in the logical language. Furthermore, Axiom 2 is then needed only once.

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly and informally described by Anderson \cite[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may only wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, if this suggests that there might have b!
 een a change (perhaps evolution) in the way logicians construct formal proofs; or even from a philosophical point of view, if G\"odel's preference was philosophically motivated.

@@

\setcounter{theorem}{2}
\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

Since the iteration principle requires the strong modal logic {\bf S5} while Brouwer's reduction principle relies only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{iteration principle}.
Anderson \cite[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep{sobel2}[p. 152] acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in Section \ref{sec:newproof} relies on {\bf KB}, which is weaker than {\bf B}.

@@

If G\"odel's ontological proof is abstractly analyzed, then we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.

@@

\bibitem[{\itshape Church (1940)}]{church}
Church, A. 1940. {\itshape A Formulation of the Simple Theory of Types}, Journal of Symbolic Logic, 5: 56--68.

@@

\bibitem[{\itshape Hajek (1996)}]{hajek}
Hajek, P. 1996. Magari and others on G\"odel's Ontological Proof, in Ursini et alii
(eds.) Logic and Logical Algebra, pp. 125--136, Marcel Dekker.

@@

\bibitem[{\itshape Sobel (1987)}]{sobel}
Sobel, J. H. 1987. {\itshape G\"odel's Ontological Proof}. In: J. J. Thompson (ed.). {\itshape On being and saying: essays for Richard Cartwright},  MIT Press.
